---
title: "GWAS"
date: "February 19, 2018"
author: "Gota Morota"
output:
   rmdformats::material:
      highlight: kate
      self_contained: no
      code_folding: hide
---


# Rice data
Download the Rice Diversity Panel data `RiceDiversity.44K.MSU6.Genotypes_PLINK.zip` from [http://ricediversity.org/data/sets/44kgwas/](http://ricediversity.org/data/sets/44kgwas/). 

## Genotype data
The genotype data set is provided in the PLINK format. We will use the `read_ped` function from the `BGLR` package to read the PLINK format files into the R environment. 
```{r eval=FALSE, echo=TRUE}
# install.packages("BGLR")
library(BGLR)
out<- read_ped("../../STAT892/data/Genotypes/sativas413.ped")
p=out$p
n=out$n
out=out$x
#Recode snp to 0,1,2 format using allele 1
# 0 --> 0
# 1 --> 1
# 2 --> NA
# 3 --> 2
out[out==2]=NA
out[out==3]=2
W <- matrix(out, nrow=p, ncol=n, byrow=TRUE)
W <- t(W) 
dim(W) # # 413 x 36901
```

## Accession IDs 
The accession IDs (individual IDs) are included in `.fam` file. 
```{r eval=FALSE}
# accession ID
fam <-read.table("../../STAT892/data/Genotypes/sativas413.fam", header = FALSE, stringsAsFactors = FALSE)  
head(fam)
rownames(W) <- fam$V2 # 413 x 36901
```

## Phenotype data
We will use the `read.table` function to read the phenotype file `RiceDiversity_44K_Phenotypes_34traits_PLINK.txt` from [here](http://www.ricediversity.org/data/sets/44kgwas/RiceDiversity_44K_Phenotypes_34traits_PLINK.txt). 
```{r eval=FALSE}
# phenotypes
rice.pheno <- read.table("http://www.ricediversity.org/data/sets/44kgwas/RiceDiversity_44K_Phenotypes_34traits_PLINK.txt", header=TRUE, stringsAsFactors = FALSE, sep = "\t")
table(rownames(W) == rice.pheno$NSFTVID)
y <- matrix(rice.pheno$Flowering.time.at.Arkansas) # # use the first trait 
rownames(y) <- rice.pheno$NSFTVID
index <- !is.na(y)
y <- y[index, 1, drop=FALSE] # 374
W <- W[index, ] # 374 x 36901
table(rownames(W) == rownames(y))
```


# Population structure
This data set is known to exhibit a subpopulation structure as the panel contains indica, aus, temperate japonica, tropical japonica, aromatic, and highly admixed accessions. We will create a PC plot to visualize the extent of population structure in rice. 
```{r eval=FALSE}
# PC plots
gp <-read.csv("http://ricediversity.org/data/sets/44kgwas/RiceDiversity.44K.germplasm.csv", header = TRUE, skip = 1,  stringsAsFactors = FALSE)   # 431 x 12
gp2 <- gp[match(rownames(y), gp$NSFTV.ID), ]
table(rownames(y) == gp2$NSFTV.ID)

	
plot(gp2$PC1, gp2$PC2, xlab="PC1", ylab="PC2", col=c(1:6)[factor(gp2$Sub.population)])
legend(x="topleft", legend = levels(factor(gp2$Sub.population)), col=c(1:6), pch=1, cex=0.6)
```



# Quality control

## Genotype imputation
Replace missing marker genotypes with mean values. Then store the marker genotypes in a matrix object `W`. 
```{r eval=FALSE,  echo=TRUE}
for (j in 1:ncol(W)){
  W[,j] <- ifelse(is.na(W[,j]), mean(W[,j], na.rm=TRUE), W[,j])
}
```


## Compute allele frequencies for all SNPs
Recall that allele frequency of *A* is given by 
$$ 
f(A) = p = \frac{2 \times (\text{no. of } AA \text{ individuals}) + 1 \times (\text{no. of } Aa \text{ individuals})}{2 \times \text{total no. of individuals}}.
$$
We can rewrite this equation into 
$$ 
f(A) = p = \frac{(\text{no. of } A \text{ allele in the population})}{2 \times \text{total no. of individuals}}.
$$
This suggests that all we need is the number of $A$ allele or reference allele $a$ for each SNP. The `sum` function returns the number of reference allele $A$. 
```{r, eval=FALSE}
sum(W[,1]) # sum of A allele in the first SNP
sum(W[,2]) # sum of A allele in the second SNP
```

How to repeat this operation for 3,6901 SNPs? The `colSums` function returns the sum of each column in a matrix as a vector. 
```{r, eval=FALSE}
colSums(W) 
```

Note that `colSums(W) ` gives the numerator of the above equation. We then divide this number by $2 \times \text{total no. of individuals}$. The function `nrows` returns the number of rows. 
```{r, eval=FALSE}
p <- colSums(W) / (2 * nrow(W)) # or colMean(X) / 2
```
The variable `p` is a vector and it contains the allele frequencies of reference allele. 


## Minor allele frequency (MAF)
Perform a quality control analysis by removing markers with MAF < 0.05. How many markers are removed? Save the filtered genotype matrix in `W2`. 
```{r eval = FALSE}
maf <- ifelse(p > 0.5, 1-p, p)
maf.index <- which(maf < 0.05)
W2 <- W[, -maf.index]  # 374 x 33558 
```


# Genome-wide association studies
Fit a single-marker-based linear mixed model  by using the `GWAS` function in the [rrBLUP](https://cran.r-project.org/web/packages/rrBLUP/index.html) R package. Report the -log10 of p-values for SNP effects. 
```{r eval=FALSE, echo=TRUE}
# install.packages("rrBLUP")
library(rrBLUP)
map <- read.table("../../STAT892/data/Genotypes/sativas413.map", header = FALSE, stringsAsFactors = FALSE)
my.geno <- data.frame(marker=map[,2], chrom=map[,1], pos=map[,4], t(W-1), check.names = FALSE) # W = \in{-1, 0, 1}
my.pheno <- data.frame(NSFTV_ID=rownames(y), y=y) 

rel <- GWAS(my.pheno, my.geno, min.MAF=0.05, P3D=TRUE, plot=FALSE)
head(rel$y)
tail(rel$y)
```


## Manhattan plot
```{r eval=FALSE, echo=TRUE}
# install.packages("qqman")
library(qqman)
manhattan(x = rel, chr = "chrom", bp = "pos", p = "y", snp = "marker", col = c("blue4", "orange3"), logp = FALSE)
```









